Comparing variations in sediment transport capacity between subglacial and subaerial systems

### 2023-03-20 theory.tex
draft to GK

  As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
  These differences have been implicitly included in a wide range of available models that  quantify sediment transport from these catchments in both subglacial and subaerial systems  \citep[e.g.][]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
  The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016}.
  Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \citep[e.g.][]{muller1968,richards2003,ganti2016}.
  Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.

@@

  Across the accepted model runs, variability in water velocity and shear stress remains substantially higher across the subglacial model compared to subaerial ones (Figure~\ref{fig:range}).
  Water velocity, shear stress, and width-integrated shear stress all have a variability ratio ($\frac{\chi(\Gamma_{g})}{\chi(\Gamma_f)}$) greater than $1$ in the subglacial  system compared to subaerial one (Figure~S1).

@@

  Runs with the smallest variability in discharge ($\chi(Q_w)$) have the greatest variability in water velocity, shear stress, and width-integrated shear stress in subglacial systems compared to subaerial ones (Figure~\ref{fig:range} and S1).
  Yet, the greatest difference in the two systems at the smallest variations of water discharge still results in greater variability in the subglacial system compared to the subaerial one (Figure \ref{fig:range}).

@@

  \subsection{Geomorphic Implications}
  \label{sect:GI}
  The relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau_t} and \ref{eq:v}).
  Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
  Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citet{meyer1948}.
  In the total sediment transport relationship \citet{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
  In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and width-integrated shear stress described above in subglacial environments (Figure~\ref{fig:model_outs}\, b and d; Figure~S1).

@@

  In large catchments and further downstream of proglacial margins, water discharge is generally less variable than at the tops of catchments, where glaciers lie in alpine settings \citep[c.f.][]{swift2005,riihimaki2005,costa2017}.
  Here, external hydrological factors further downstream of glaciers could result in smaller variations in sediment transport capacity in many subaerial systems compared to subglacial ones.

  The parameterization here shows that the smallest variability in water discharge has the largest difference in relative variability in sediment transport parameters (Figure~\ref{fig:range}\, d,\,e,\,f).
  At locations such as the Greenland Ice Sheet, smaller diurnal variations in water discharge compared to many alpine catchments \citep[c.f.][]{swift2005,riihimaki2005,vanas2017,hasholt2018}.
  Therefore, the variability in sediment transport capacity in the transition from the ice sheet to the proglacial river could be more dramatic in Greenland than in alpine systems as water leaves the glacier and moves downstream.
  Thus the disparity in sediment transport capacity between the subglacial and subaerial systems could be most pronounced in larger catchments, potentially resulting in temporally mismatched periods between when glaciers introduce sediment to these river systems  and when the river mobilizes it (Figures~\ref{fig:model_outs}~ and~\ref{fig:range}).

@@

>>>>

  As a result, sediment mobilization in subaerial and subglacial systems responds differently to changing water discharge.
  These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial systems  \citep[e.g.][]{walder1994,tucker1997,creyts2013,wickert2019,hewitt2019}.
  The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016}.
  Additionally, the variable response of sediment transport capacity to water discharge in the two systems may impact the interpretation of records of sediment transport in glacial compared to subaerial systems \citep[e.g.][]{muller1968,richards2003,ganti2016}.
  Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial systems has been minimally discussed.

@@

  Across the accepted model runs, variability in water velocity and shear stress remains substantially higher across the subglacial model compared to subaerial ones (Figure~\ref{fig:range}).
  Water velocity, shear stress, and width-integrated shear stress all have a variability ratio ($\frac{\chi(\Gamma_{g})}{\chi(\Gamma_f)}$) greater than $1$ in the subglacial  system compared to subaerial one (Figure~S \ref{fig:gammas}).

@@

  Runs with the smallest variability in discharge ($\chi(Q_w)$) have the greatest variability in water velocity, shear stress, and width-integrated shear stress in subglacial systems compared to subaerial ones (Figure~\ref{fig:range} and S \ref{fig:gammas}).
  Yet, the greatest difference in the two systems at the smallest variations of water discharge still results in greater variability in the subglacial system compared to the subaerial one (Figure \ref{fig:range}).

@@

  \subsection{Geomorphic Implications}
  \label{sect:GI}
  The relationship between water discharge variability and relative variability of the water velocity, shear stress, and width-integrated shear stress results from the fixed geometry in the glacier system and variable geometry in the subaerial system as they respond to changes in water discharge (Equations~\ref{eq:tau_t} and \ref{eq:v}).
  Increased variability in sediment transport capacity with respect to discharge has several implications for sediment transport in subglacial systems compared to subaerial ones.
  Shear stress is scaled to the $\frac{3}{2}$ power and relies on shear stress exceeding a threshold in sediment transport relationships  such as in \citet{meyer1948}.
  In the total sediment transport relationship \citet{engelund1967}, the shear stress is scaled to $\frac{5}{2}$ power.
  In both cases, the exponent greater than $1$ of shear stress in the sediment transport relationship magnifies the variability in sediment transport beyond the highly variable shear stress and width-integrated shear stress described above in subglacial environments (Figure~\ref{fig:model_outs}\, b and d; Figure~S~\ref{fig:gammas}).

@@

  In large catchments and further downstream of proglacial margins, water discharge is generally less variable than at the tops of catchments, where glaciers lie in alpine settings \citep[c.f.][]{swift2005,riihimaki2005,costa2017,vanas2017}.
  Here, external hydrological factors further downstream of glaciers could result in smaller variations in sediment transport capacity in many subaerial systems compared to subglacial ones.

  The parameterization here shows that the smallest variability in water discharge has the largest difference in relative variability in sediment transport metrics (Figure~\ref{fig:range}\, d,\,e,\,f).
  At locations such as the Greenland Ice Sheet, smaller diurnal variations in water discharge compared to many alpine catchments \citep[c.f.][]{swift2005,riihimaki2005,vanas2017,hasholt2018}.
  Therefore, the variability in sediment transport capacity in the transition from the ice sheet to the proglacial river could be more dramatic in Greenland than in alpine systems as water leaves the glacier and moves downstream.
  Thus the disparity in sediment transport capacity between the subglacial and subaerial systems could be most pronounced in larger catchments, potentially resulting in temporally mismatched periods between when glaciers introduce sediment to these river systems  and when the river mobilizes it (Figures~\ref{fig:model_outs}~ and~\ref{fig:range}).

@@

\renewcommand{\figurename}{Figure S}
\setcounter{figure}{0}
